Inference using context-dependent models of mutation via bounds on information propagation

### 2021-12-10 writeups/context-dependence.tex
Finishing my pass through main text

%EM It seems reasonable that we should be able to get away using a smaller T-mer.
%EM Question: how much context should we include; argument in appendix.
%EM Would be nice to show convergence of full likelihood eventually.

@@

%%%%%%%%
\subsection{Model fit and model selection}

Fitting a model to nucleotide data would likely require iterative modification of the model.
The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
For each T-mer $(a,b)$, write $O_{a,b}$ for the number of this T-mer observed in the data,
and $E_{a,b}$ for the number expected under the model.
(The expected number is the number of occurrences of $a$ in the initial sequence
multiplied by the probability $p_{n,\ell,r}(t;a,b)$, where $n$ is the length of $a$,
and $\ell$, $r$ are the appropriate overhangs.)
If the T-mers were nonoverlapping, this would form a standard contingency table,
and standard practice would be to divide the residuals by the square root of the expected counts
and compare these to a standard Normal distribution.
However, there are are $n$ possible such tables --
so, we compute normalized residuals as
\begin{align}
    Z_{a,b} = \frac{ O_{a,b} - E_{a,b} }{ \sqrt{n E_{a,b}} },
\end{align}
which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
These can then be compared to a standard Normal distribution.

@@

The general case is a modification of Felsenstein's ``pruning'' algorithm.

@@

In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
However, we can reduce dimensionality by only computing probabilities
for certain (more common) patterns: for instance,
we'd compute the likelihood for situations where $y$ and $z$ agree (but may differ from $x$
by replacing step (4) with:
\begin{enumerate}

  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{n}} p_{n,0,0}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{n} \times |\calS|^{m}$ matrix).

@@

%%%%%%%%%%%
\subsection{Hominid data}
\label{hominid_section}

We also applied the model to human--chimpanzee divergence data,
restricting the comparison to putative regulatory sequences,
avoiding the additional constraints of coding sequences.
To do this, we extracted regions in the human (hg38)--chimpanzee (panTro4) alignment % \citep[from the UCSC Genome Browser][]{pantroalign}
according to four types of regulatory feature in the Ensembl Regulatory Build \citep[release 81][]{zerbino2015ensembl}:
``CTCF binding site'', ``enhancer'', ``open chromatin region'', and ``promoter flanking region''.
These were furthermore divided into regions that either did or did not overlap a transcription start or end site of a known gene
from the hg38 Annotations Database \citep{kent2002human}
and filtered for length:
aligned regions below either 3000bp (promoter flanking regions) or 1000bp (all other types).
We then counted T-mers of appropriate length in each of the eight resulting sets of aligned regions, omitting any T-mers that included gaps or repeat-masked positions,
and fit models separately to each of the eight data sets.

@@

      \item[(repair)] In addition to the above, we included a collection of mutation motifs associated with known DNA lesion/repair pathways
          curated from the literature \citep[reviewed in][]{sale2012yfamily,goodman2013translesion,roberts2014hypermutation,cobey2015evolution,goodman2002errorprone}:
          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$,][]{sale2012yfamily,sinha2002uvinduced};
          repair of guanine adducts at CpG sites \citet[$\mu_g$,][]{pfeifer2006mutagenesis};
          deamination by AID \citep[$\mu_{AID}$,][]{teng2007immunoglobulin,kasar2015wholegenome};
          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$,][]{alexandrov2013deciphering};
          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$,][]{alexandrov2013deciphering};
          and pol $\iota$ \citep[$\mu_{\iota}$,][]{roberts2014hypermutation,maul2016polymerase}.
          These are denoted here using ambiguity codes
          \nR{}: \nA{} or \nG{},
          \nY{}: \nC{} or \nT{},
          \nW{}: \nA{} or \nT{}, and
          \nS{}: \nC{} or \nG{};
          so for instance
          ``\nT\nC\nW$\to$\nT\nT\nW''
          means that both
          \nT\nC\nA$\to$\nT\nT\nA{}
          and
          \nT\nC\nT$\to$\nT\nT\nT{}
          share the same mutation rate.
          \noindent
      \begin{center}
          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
            \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA       &   \nA\nG       &  $\mu_{\iota1}$   \\
            \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nT\nT       &   \nC\nT       &  $\mu_{\iota1}$   \\
            \nT\nC     &   \nT\nT      &  $\mu_{UV}$     & \nA\nA       &   \nG\nT       &  $\mu_{\iota2}$   \\
            \nG\nA     &   \nA\nA      &  $\mu_{UV}$     & \nT\nT       &   \nA\nC       &  $\mu_{\iota2}$   \\
            \nC\nC     &   \nC\nT      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nT\nW    &  $\mu_{\APOBEC}$  \\
            \nG\nG     &   \nA\nG      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nG\nW    &  $\mu_{\APOBEC}$  \\
            \nC\nC     &   \nT\nT      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nA\nA    &  $\mu_{\APOBEC}$  \\
            \nG\nG     &   \nA\nA      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nC\nA    &  $\mu_{\APOBEC}$  \\
            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nW\nR\nC\nY &  \nW\nR\nT\nY  &  $\mu_{\AID}$     \\
            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nR\nG\nY\nS &  \nR\nA\nY\nS  &  $\mu_{\AID}$
        \end{tabular}
      \end{center}

@@

In exploratory application to real data, how does one decide when to stop adding mutational motifs?
Perhaps the simplest method conceptually is to
(a) fit a simple model,
(b) choose new motifs to add based on statistically significant residuals,
(c) repeat until no residuals are statistically significant.
To test this method,
we simulated a single sequence of $10^6$ nucleotides
for one unit of time
under the model given in Table \ref{tab:cpg_results},
obtained by embellishing the basic CpG hypermutability model
with the longer mutational motif identified by \citet{harris2015evidence}.
%     \begin{center}
%       \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
%         $x$  &    $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
%         \nC\nG     &  \nT\nG     &  $0.4$ & \\
%         \nC\nG     &  \nC\nA     &  $0.4$ & \\
%         \nT\nC\nC  &  \nT\nT\nC  &  $0.1$ & \\
%         \nA\nG\nG  &  \nA\nA\nG  &  $0.1$ &
%       \end{tabular} ,
%     \end{center}
% with $m_{AT}=m_{TA}=0.1$, $m_{CG}=m_{GC}=0.15$, $m_{AC}=m_{TG}=m_{AG}=m_{TC}=0.08$, and $m_{CA}=m_{GT}=m_{CT}=m_{GA}=0.12$.
% Note the model is symmetric under reverse complementation.
The resulting sequences differed at 28.5\% of the sites.
We then fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
(three sites at the base with an overhang of one on each side)
using a constrained quasi-Newton method (``optim(.., method='L-BFGS-B')'' in R).
The top and bottom of the table of (2,1,0) and (2,0,1) T-mer residuals of this model were
as shown in Table~\ref{tab:resid1}.

\begin{table}
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & MLE \\
          \hline
          \nA  &   \nT           & 0.10 & 0.0996 \\
          \nT  &   \nA           & 0.10 & 0.0982 \\
          \nC  &   \nG           & 0.15 & 0.1505 \\
          \nG  &   \nC           & 0.15 & 0.1497 \\
          \nA  &   \nC           & 0.08 & 0.0784 \\
          \nT  &   \nG           & 0.08 & 0.0792 \\
          \nA  &   \nG           & 0.08 & 0.0798 \\
          \nT  &   \nC           & 0.08 & 0.0795 \\
          \nC  &   \nA           & 0.12 & 0.1202 \\
          \nG  &   \nT           & 0.12 & 0.1196 \\
          \nC  &   \nT           & 0.12 & 0.1198 \\
          \nG  &   \nA           & 0.12 & 0.1193 \\
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
           \hline
        \end{tabular}
  \end{center}
  \caption{
      True parameter values, and estimates obtained
      from iterative fitting to a $10^6$ bp sequence
    evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$
    were constrained during fitting to be equal (DNA strand symmetry),
    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
    \label{tab:cpg_results} }
\end{table}

\begin{table}
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l1.tsv
    \begin{center}
        \begin{tabular}{ccrrrr}
                $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
                \hline
                \nC\nG  & \nC\_   & 100410   &  120303    & -19892.50   &	-40.55  \\
                \nC\nA  & \nT\_   &  19884   &   28350    &  -8466.43   &	-35.55  \\
                \nC\nT  & \nT\_   &  19998   &   28061    &  -8063.46   &	-34.03  \\
                \nC\nC  & \nT\_   &  23673   &   28177    &  -4503.96   &	-18.97  \\
                \nG\nG  & \nC\_   &  17742   &   19418    &  -1675.92   &	 -8.50  \\
                \nC\nG  & \nG\_   &  18450   &   19478    &  -1028.34   &	 -5.21  \\
                \nA\nG  & \nC\_   &  10968   &   11760    &   -791.73   &	 -5.16  \\
                \nG\nG  & \nG\_   & 118805   &  120975    &  -2170.25   &    \\
                \hline
                \nA\nG  & \nT\_   &  17280   &   16515    &    765.20   &  4.21   \\
                \nG\nT  & \nC\_   &  20490   &   19507    &    983.44   &  4.97   \\
                \nG\nG  & \nT\_   &  20929   &   19278    &   1650.83   &  8.40   \\
                \nC\nC  & \nC\_   & 124149   &  119880    &   4268.52   &  8.71   \\
                \nG\nG  & \nA\_   &  30483   &   28288    &   2195.34   &  9.22   \\
                \nC\nT  & \nC\_   & 126957   &  119389    &   7567.92   & 15.48   \\
                \nC\nA  & \nC\_   & 128673   &  120619    &   8054.46   & 16.39   \\
                \nC\nG  & \nT\_   &  49311   &   28276    &  21034.85   & 88.45   \\
                \hline
                \hline
                \nC\nG  &  \_\nG  &  100803  &  120541  &  -19737.80  &  -40.19 \\
                \nT\nG  &  \_\nA  &   20040  &   28179  &   -8139.45  &  -34.28 \\
                \nG\nG  &  \_\nA  &   20985  &   28288  &   -7302.96  &  -30.70 \\
                \nA\nG  &  \_\nA  &   22929  &   28399  &   -5469.73  &  -22.95 \\
                \nC\nC  &  \_\nG  &   17805  &   19410  &   -1605.01  &   -8.14 \\
                \nC\nT  &  \_\nG  &   10470  &   11684  &   -1214.31  &   -7.94 \\
                \nC\nA  &  \_\nG  &   10839  &   11882  &   -1043.39  &   -6.76 \\
                \nC\nG  &  \_\nC  &   18354  &   19348  &    -994.19  &   -5.05 \\
                \hline
                \nC\nT  &  \_\nA  &   17007  &   16078  &     929.33  &    5.18 \\
                \nT\nC  &  \_\nG  &   20691  &   19509  &    1182.40  &    5.98 \\
                \nC\nC  &  \_\nA  &   20640  &   19160  &    1480.44  &    7.56 \\
                \nT\nC  &  \_\nT  &   30309  &   28320  &    1988.91  &    8.35 \\
                \nA\nG  &  \_\nG  &  127191  &  121450  &    5740.76  &   11.64 \\
                \nG\nG  &  \_\nG  &  127946  &  120977  &    6969.46  &   14.16 \\
                \nT\nG  &  \_\nG  &  127539  &  120512  &    7026.52  &   14.31 \\
                \nC\nG  &  \_\nA  &   49098  &   28186  &   20911.93  &   88.07 \\
                \hline
        \end{tabular}
    \end{center}
    \caption{
        Largest residuals of shape (2,1,0) and (2,0,1) after fitting a model with only ``basic''
        (single-nucleotide) mutations to sequence simulated under the model
        of Table~\ref{tab:cpg_results}, that also includes CpG hypermutability
        as well as $TCC \to TTC$ and $AGG \to AAG$ mutations.
        Note that in the table, the underscore means ``any nucleotide''.
        \label{tab:resid1}
    }
\end{table}

The situation clearly calls for addition of the CpG mutation motif.
After adding these patterns into the model and re-fitting,
% starting at the parameters inferred from the first run and a CpG rate of 0.0,
some length 2 residuals had statistically significant residuals,
but none so large as those seen in (3,1,1) T-mer residuals, shown in Table~\ref{tab:resid2}.

>>>>

@@

@@

The general case is a modification of Felsenstein's ``pruning'' algorithm\citep{felsenstein1981evolutionary}.

@@

In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
However, we can reduce dimensionality by only computing probabilities
for certain (more common) patterns: for instance,
one could compute the likelihood for situations where $y$ and $z$ agree (but may differ from $x$
by replacing step (4) with:
\begin{enumerate}

  \item[$4'$.] Compute $M_4(x,w) = \sum_{v \in \calS^{n}} p_{n,0,0}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{n} \times |\calS|^{m}$ matrix).

@@

%%%%%%%%
\subsection{Model fit and model selection}
%EM I promoted this section here because we use residuals a lot in the results.
Fitting a model to nucleotide data will likely require iterative modification of the model.
The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
%EM Why shorter? Because there are fewer of them?
These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
%EM Should we replace a and b with x and y here to agree with the rest of the doc?
For each T-mer $(a,b)$, write $O_{a,b}$ for the number of this T-mer observed in the data,
and $E_{a,b}$ for the number expected under the model.
(The expected number is the number of occurrences of $a$ in the initial sequence
multiplied by the probability $p_{n,\ell,r}(t;a,b)$, where $n$ is the length of $a$,
and $\ell$, $r$ are the appropriate overhangs.)
If the T-mers were nonoverlapping, this would form a standard contingency table,
and standard practice would be to divide the residuals by the square root of the expected counts
and compare these to a standard Normal distribution.
However, there are are $n$ possible such tables --
so, we compute normalized residuals as
\begin{align}
    Z_{a,b} = \frac{ O_{a,b} - E_{a,b} }{ \sqrt{n E_{a,b}} },
\end{align}
which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
These can then be compared to a standard Normal distribution.

We benchmarked the performance of stepwise model selection:
(a) fit a simple model,
(b) choose new motifs to add based on statistically significant residuals,
(c) repeat until no residuals are statistically significant.
To test this method,
we simulated a single sequence of $10^6$ nucleotides
for one unit of time
under the model given in Table \ref{tab:cpg_results},
obtained by embellishing the basic CpG hypermutability model
with the longer mutational motif identified by \citet{harris2015evidence}.
%     \begin{center}
%       \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
%         $x$  &    $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
%         \nC\nG     &  \nT\nG     &  $0.4$ & \\
%         \nC\nG     &  \nC\nA     &  $0.4$ & \\
%         \nT\nC\nC  &  \nT\nT\nC  &  $0.1$ & \\
%         \nA\nG\nG  &  \nA\nA\nG  &  $0.1$ &
%       \end{tabular} ,
%     \end{center}
% with $m_{AT}=m_{TA}=0.1$, $m_{CG}=m_{GC}=0.15$, $m_{AC}=m_{TG}=m_{AG}=m_{TC}=0.08$, and $m_{CA}=m_{GT}=m_{CT}=m_{GA}=0.12$.
% Note the model is symmetric under reverse complementation.
The resulting sequences differed at 28.5\% of the sites.

To begin, we fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
(three sites at the base with an overhang of one on each side)
using a constrained quasi-Newton method (\verb|optim(.., method='L-BFGS-B')| in R).

\begin{table}[ht]
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & MLE \\
          \hline
          %EM I'm confused why we have MLE parameters for these single NT terms because I don't see those being used in the iterative model fitting. I thought we started with a (5,1,1) model.
          \nA  &   \nT           & 0.10 & 0.0996 \\
          \nT  &   \nA           & 0.10 & 0.0982 \\
          \nC  &   \nG           & 0.15 & 0.1505 \\
          \nG  &   \nC           & 0.15 & 0.1497 \\
          \nA  &   \nC           & 0.08 & 0.0784 \\
          \nT  &   \nG           & 0.08 & 0.0792 \\
          \nA  &   \nG           & 0.08 & 0.0798 \\
          \nT  &   \nC           & 0.08 & 0.0795 \\
          \nC  &   \nA           & 0.12 & 0.1202 \\
          \nG  &   \nT           & 0.12 & 0.1196 \\
          \nC  &   \nT           & 0.12 & 0.1198 \\
          \nG  &   \nA           & 0.12 & 0.1193 \\
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
           \hline
        \end{tabular}
  \end{center}
  \caption{
      True parameter values, and MLE estimates obtained
      from iterative fitting as described in the text to a $10^6$ bp sequence
    evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $\nC\nG \to \nT\nG$ and $\nC\nG \to \nC\nA$
    were constrained during fitting to be equal (DNA strand symmetry),
    as were the transitions $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
    \label{tab:cpg_results} }
\end{table}

We find that the (2,1,0) and (2,0,1) T-mer residuals of this model are dominated by
terms involving CG (Table~\ref{tab:resid1}).
%EM Stupid question: why weren't these picked up by the 5,1,1 model?
%EM Why isn't the starting point a (1,0,0) model?
These results clearly call for addition of the CpG mutation motif.

\begin{table}
%EM This table is BIG. And it seems like our conclusions only stem from the very biggest residuals.
%EM What if we cut of anything that doesn't have an absolute Z-score bigger than 10?
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l1.tsv
    \begin{center}
        \begin{tabular}{ccrrrr}
                $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
                \hline
                \nC\nG  & \nC\_   & 100410   &  120303    & -19892.50   &	-40.55  \\
                \nC\nA  & \nT\_   &  19884   &   28350    &  -8466.43   &	-35.55  \\
                \nC\nT  & \nT\_   &  19998   &   28061    &  -8063.46   &	-34.03  \\
                \nC\nC  & \nT\_   &  23673   &   28177    &  -4503.96   &	-18.97  \\
                \nG\nG  & \nC\_   &  17742   &   19418    &  -1675.92   &	 -8.50  \\
                \nC\nG  & \nG\_   &  18450   &   19478    &  -1028.34   &	 -5.21  \\
                \nA\nG  & \nC\_   &  10968   &   11760    &   -791.73   &	 -5.16  \\
                %EM It looks like this is missing an entry?
                \nG\nG  & \nG\_   & 118805   &  120975    &  -2170.25   &    \\
                \hline
                \nA\nG  & \nT\_   &  17280   &   16515    &    765.20   &  4.21   \\
                \nG\nT  & \nC\_   &  20490   &   19507    &    983.44   &  4.97   \\
                \nG\nG  & \nT\_   &  20929   &   19278    &   1650.83   &  8.40   \\
                \nC\nC  & \nC\_   & 124149   &  119880    &   4268.52   &  8.71   \\
                \nG\nG  & \nA\_   &  30483   &   28288    &   2195.34   &  9.22   \\
                \nC\nT  & \nC\_   & 126957   &  119389    &   7567.92   & 15.48   \\
                \nC\nA  & \nC\_   & 128673   &  120619    &   8054.46   & 16.39   \\
                \nC\nG  & \nT\_   &  49311   &   28276    &  21034.85   & 88.45   \\
                \hline
                \hline
                \nC\nG  &  \_\nG  &  100803  &  120541  &  -19737.80  &  -40.19 \\
                \nT\nG  &  \_\nA  &   20040  &   28179  &   -8139.45  &  -34.28 \\
                \nG\nG  &  \_\nA  &   20985  &   28288  &   -7302.96  &  -30.70 \\
                \nA\nG  &  \_\nA  &   22929  &   28399  &   -5469.73  &  -22.95 \\
                \nC\nC  &  \_\nG  &   17805  &   19410  &   -1605.01  &   -8.14 \\
                \nC\nT  &  \_\nG  &   10470  &   11684  &   -1214.31  &   -7.94 \\
                \nC\nA  &  \_\nG  &   10839  &   11882  &   -1043.39  &   -6.76 \\
                \nC\nG  &  \_\nC  &   18354  &   19348  &    -994.19  &   -5.05 \\
                \hline
                \nC\nT  &  \_\nA  &   17007  &   16078  &     929.33  &    5.18 \\
                \nT\nC  &  \_\nG  &   20691  &   19509  &    1182.40  &    5.98 \\
                \nC\nC  &  \_\nA  &   20640  &   19160  &    1480.44  &    7.56 \\
                \nT\nC  &  \_\nT  &   30309  &   28320  &    1988.91  &    8.35 \\
                \nA\nG  &  \_\nG  &  127191  &  121450  &    5740.76  &   11.64 \\
                \nG\nG  &  \_\nG  &  127946  &  120977  &    6969.46  &   14.16 \\
                \nT\nG  &  \_\nG  &  127539  &  120512  &    7026.52  &   14.31 \\
                \nC\nG  &  \_\nA  &   49098  &   28186  &   20911.93  &   88.07 \\
                \hline
        \end{tabular}
    \end{center}
    \caption{
        Largest residuals of shape (2,1,0) and (2,0,1) after fitting a model with only ``basic''
        (single-nucleotide) mutations to sequence simulated under the model
        of Table~\ref{tab:cpg_results}, that also includes CpG hypermutability
        as well as $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ mutations.
        Note that in the table, the underscore means ``any nucleotide''.
        \label{tab:resid1}
    }
\end{table}

%EM "These patterns": Is it exactly the patterns in the true model?
%EM Would we need to have prior information about CpGs to pick those additional terms?
After adding these patterns into the model and re-fitting, we searched for evidence of additional processes happening in the data.
% starting at the parameters inferred from the first run and a CpG rate of 0.0,
%EM what is "statistically significant" here? I searched around and found 1.96 corresponding to 0.05.
To do so, we calculated length 2 residuals and (3,1,1) T-mer residuals.
Although some of the length 2 residuals were statistically significant,
none were as large as those seen in (3,1,1) T-mer residuals, which are shown in Table~\ref{tab:resid2}.
These residuals clearly call for adding the mutation tuples $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
These were indeed the motifs used when simulating the data.
After adding these in,
% and initializing all parameters to 0.1
no further (3,1,1) T-mer residuals achieved
statistical significance after a Bonferroni correction;
the final parameter values are shown in Table \ref{tab:cpg_results};
parameter estimates differ from the truth by less than 3\% (mostly within 1\%).

\begin{table}
    \begin{center}
        \begin{tabular}{ccrrrr}
            \hline
                $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
                \hline
                \nA\nG\nG  &  \_\nG\_  &  30024  &  32167  &  -2143.48  &  -6.90 \\
                \nT\nC\nC  &  \_\nC\_  &  29910  &  31748  &  -1838.13  &  -5.95 \\
                \nG\nG\nT  &  \_\nA\_  &   5064  &   5549  &   -484.61  &  -3.75 \\
                \nT\nG\nC  &  \_\nA\_  &   4830  &   5172  &   -342.42  &  -2.74 \\
                \nG\nG\nA  &  \_\nA\_  &   5115  &   5462  &   -346.97  &  -2.71 \\
                \nC\nT\nT  &  \_\nG\_  &   2472  &   2717  &   -244.56  &  -2.70 \\
                \nG\nC\nT  &  \_\nT\_  &   4863  &   5192  &   -328.56  &  -2.63 \\
                \hline
                \nA\nT\nG  &  \_\nG\_  &   3225  &   3028  &    196.78  &   2.06  \\
                \nA\nC\nG  &  \_\nA\_  &   4830  &   4588  &    242.29  &   2.06  \\
                \nT\nG\nT  &  \_\nC\_  &   5364  &   5105  &    258.74  &   2.09  \\
                \nG\nA\nG  &  \_\nG\_  &   3285  &   3055  &    229.78  &   2.40  \\
                \nT\nT\nG  &  \_\nC\_  &   2973  &   2752  &    220.59  &   2.42  \\
                \nT\nC\nT  &  \_\nG\_  &   5421  &   5056  &    364.54  &   2.95  \\
                \nT\nC\nC  &  \_\nT\_  &   7590  &   5500  &   2089.79  &  16.26  \\
                \nA\nG\nG  &  \_\nA\_  &   7566  &   5263  &   2302.74  &  18.32  \\
                \hline
        \end{tabular}
    \end{center}
    \caption{
        Largest residuals of shape (3,1,1) after fitting a model with ``basic''
        (single-nucleotide) and CpG mutations to sequence simulated under the model
        of Table~\ref{tab:cpg_results}, that also includes $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ mutations.
        Note that in the table, the underscore means ``any nucleotide''.
        \label{tab:resid2}
    }
\end{table}

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hastings-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
for 1 unit of time,
resulting in around 460,000 possible mutations
and differences at around 186,000 sites between initial and final sequence.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
counted $(5,3,1)$ T-mers,
and used the \texttt{mcmc} package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
(judged sufficient by examination of diagnostic plots).
We repeated this procedure separately on 61 independently simulated sequences,
and show the resulting credible intervals in Figure~\ref{fig:ising_coverage}.

\begin{figure}
    \begin{center}
        %EM Should I be looking at the gray bars or black bars? What are they? I'm guessing 95% is gray, because that corresponds to good coverage.
        \includegraphics{writeup-plots/coverage_results}
    \end{center}
    \caption{
        Credible intervals for the four parameters of the dynamic Ising model,
        calculated using $(6,2,2)$ T-mers,
        from 61 MCMC runs on independently simulated sequences of $10^6$ bases
        differing at around 18.6\% of the sites.
        Shown are relative intervals, obtained by dividing by the true value.
        \label{fig:ising_coverage}}
\end{figure}

We also applied the method to human--chimpanzee divergence data,
restricting the comparison to putative regulatory sequences,
avoiding the additional constraints of coding sequences.
To do this, we extracted regions in the human (hg38)--chimpanzee (panTro4) alignment % \citep[from the UCSC Genome Browser][]{pantroalign}
according to four types of regulatory feature in the Ensembl Regulatory Build \citep[release 81][]{zerbino2015ensembl}:
``CTCF binding site'', ``enhancer'', ``open chromatin region'', and ``promoter flanking region''.
These were furthermore divided into regions that either did or did not overlap a transcription start or end site of a known gene
from the hg38 Annotations Database \citep{kent2002human}
and filtered for length:
aligned regions below either 3000bp (promoter flanking regions) or 1000bp (all other types).
We then counted T-mers of appropriate length in each of the eight resulting sets of aligned regions, omitting any T-mers that included gaps or repeat-masked positions,
and fit models separately to each of the eight data sets.

@@

      \item[(repair)] In addition to the above, we included a collection of mutation motifs associated with known DNA lesion/repair pathways
          curated from the literature \citep[reviewed in][]{sale2012yfamily,goodman2013translesion,roberts2014hypermutation,cobey2015evolution,goodman2002errorprone}:
          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$,][]{sale2012yfamily,sinha2002uvinduced};
          repair of guanine adducts at CpG sites \citet[$\mu_g$,][]{pfeifer2006mutagenesis};
          deamination by AID \citep[$\mu_\AID$,][]{teng2007immunoglobulin,kasar2015wholegenome};
          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$,][]{alexandrov2013deciphering};
          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$,][]{alexandrov2013deciphering};
          and pol $\iota$ \citep[$\mu_{\iota}$,][]{roberts2014hypermutation,maul2016polymerase}.
          These are denoted here using ambiguity codes
          \nR{}: \nA{} or \nG{},
          \nY{}: \nC{} or \nT{},
          \nW{}: \nA{} or \nT{}, and
          \nS{}: \nC{} or \nG{};
          so for instance
          ``\nT\nC\nW$\to$\nT\nT\nW''
          means that both
          \nT\nC\nA$\to$\nT\nT\nA{}
          and
          \nT\nC\nT$\to$\nT\nT\nT{}
          share the same mutation rate.
          \noindent
      \begin{center}
          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
            \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA       &   \nA\nG       &  $\mu_{\iota1}$   \\
            \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nT\nT       &   \nC\nT       &  $\mu_{\iota1}$   \\
            \nT\nC     &   \nT\nT      &  $\mu_{UV}$     & \nA\nA       &   \nG\nT       &  $\mu_{\iota2}$   \\
            \nG\nA     &   \nA\nA      &  $\mu_{UV}$     & \nT\nT       &   \nA\nC       &  $\mu_{\iota2}$   \\
            \nC\nC     &   \nC\nT      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nT\nW    &  $\mu_{\APOBEC}$  \\
            \nG\nG     &   \nA\nG      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nG\nW    &  $\mu_{\APOBEC}$  \\
            \nC\nC     &   \nT\nT      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nA\nA    &  $\mu_{\APOBEC}$  \\
            \nG\nG     &   \nA\nA      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nC\nA    &  $\mu_{\APOBEC}$  \\
            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nW\nR\nC\nY &  \nW\nR\nT\nY  &  $\mu_{\AID}$     \\
            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nR\nG\nY\nS &  \nR\nA\nY\nS  &  $\mu_{\AID}$
        \end{tabular}
      \end{center}

@@

Nevertheless, we find significant effects for a number of biochemically-informed mutation features (Figure~\ref{fig:biochem_results}).
Specifically, we find strong CpG and UV damage effects.
%EM Kelley saw the same thing https://paperpile.com/shared/aAYSgW ...
It may seem tenuous to look for mutational signatures associated with cancer in the germline,
or absurd that UV damage could play a significant role.
This is in part due to the ``streetlight effect'' --
we can't look for signatures that are not characterized --
but it is also reasonable that shared repair machinery might induce similar context-dependent errors.
More likely sources of error in the germline include the effects of recombination
\citep{arbeithuber2015crossovers,myers2010drive}.
